Reinforcement \change{L}earning for \change{O}ptions\\on \change{T}arget \change{V}olatility \change{F}unds\thanks{Supported by ISCRA from Cineca, \email{iscra@cineca.it}.}

### 2023-02-26 tex/m2ef/daluiso_nastasi_pallavicini_polo_rl_for_options_on_tvf.tex
corrections of RL sections

\usepackage[]{latexsym}
\usepackage{graphicx}
\usepackage{xcolor}

@@

\usepackage[normalem]{ulem}
\newcommand\soutpars[1]{\let\helpcmd\sout\parhelp#1\par\relax\relax}
\long\def\parhelp#1\par#2\relax{%
  \helpcmd{#1}\ifx\relax#2\else\par\parhelp#2\relax\fi%
}
\newcommand{\change}[1]{{\color{red} {#1}}}%{{#1}}
\newcommand{\remove}[1]{{\color{red} \soutpars{{#1}}}}%{}
\newcommand{\changeB}[1]{{\color{blue} {#1}}}%{{#1}}
\newcommand{\removeB}[1]{{\color{blue} \soutpars{{#1}}}}%{}

@@

In this way, the TVS is only defined in the two contracts client-fund and fund-\change{option writer} \remove{fund-bank}. The fund manager is not implementing the strategy by trading in the market, and \remove{he} \change{she} is not subject to additional costs to access the market. On the other way, the \remove{bank} \change{writer} is paying such costs since she is actively hedging the call option sold to the manager.  The \remove{bank} \change{writer} trading activity implemented to actively hedge the option requires funding the collateral procedures of the hedging instruments along with any lending/borrowing fee.
We remark that\change{, upon entering into the contract,}  the choices of the manager trading activity \remove{are} \change{can be seen as} stochastic processes since they \change{are not known {\it a priori} and }will depend on the market evolution. Thus, neither the manager strategy nor the \remove{bank} \change{writer} financing costs for hedging the option can be written in the fund-\remove{bank}\change{writer} contract\change{: the only contractual elements are the protection strike and maturity, the trading market of the manager and the target volatility of the TVS portfolio}. For this reason, the price of a financial product sold by the \remove{bank} \change{writer} is adjusted to include any valuation adjustment due to the trading activity.

Our aim is to find the most expensive investing strategy from the point of view of financing costs for the \remove{bank} \change{writer} that the manager could choose in the market. In other words, we want to determine the worst-case scenario for the bank. In the current section, we proceed by defining the price process of the TVS so that we can highlight the impact of valuation adjustments.

@@

Then, we introduce the strategy price process $V_t$, and we define the deflated gain process $\Bar{G}_t^V$ as \remove{given by}
\begin{equation}
    \Bar{G}_t^V \coloneqq \frac{V_t}{B_t} + \int_0^t\frac{V_u\phi_u}{B_u}du \; ,
\end{equation}
where $\phi_t$ are the running fees earned by the fund manager for his activity. We assume that the strategy is self-financing, so that we can write
\begin{equation}
    d\Bar{G}_t^V = q_t \cdot d\bar{G}_t \; ,
\label{eq:self_financing}\end{equation}
where $q_t^i$ is the quantity invested in the $i$-th security.\footnote{In all formulae we use dot notation for scalar product between vectors, i.e. $a \cdot b = \sum_i a_i b_i$, or between matrix and vector, i.e. $A \cdot b = \sum_j a_{ij}b_j$ or $b \cdot A = \sum_i b_i a_{ij}$.}

@@

\begin{proposition}
When the underlying securities follow a Black and Scholes \change{QUA IO LASCEREI MODEL (REFEREE PROPONE SCENARIO)} model with deterministic rates, the optimal strategy for a non-decreasing European payoff consists in minimizing the local drift function, independently of the current state $I_t$
\begin{equation}
    \alpha^*(t) \coloneqq \argmin_\alpha \frac{\alpha \cdot \mu(t)}{\|\alpha \cdot \nu(t) \|} \; .
\label{eq:BS_optimal_strategy}
\end{equation}
Analogously, the optimal strategy for a non-increasing European payoff consists in maximizing the local drift function:
\begin{equation}
    \alpha^*(t) \coloneqq \argmax_\alpha \frac{\alpha \cdot \mu(t)}{\norm{\alpha \cdot \nu(t)}} \; .
\label{eq:BS_optimal_strategy_put}
\end{equation}
\end{proposition}
The absence of stochastic elements in Equation~\eqref{eq:BS_optimal_strategy} makes the optimal strategy known \textit{a priori} with no simulation needed; in fact one can solve the optimization problem once for all $\forall t \in \mathcal{T}$ just looking at the market data $\mu(t)$ and $\nu(t)$ for the securities. Once $\alpha^*$ is known, then one can price the payoff by the following BS formula
\begin{equation}
	V_0^{\text{BS}} = BS(F^{\text{TVS}}(0,T;\alpha^*), K, T, \bar{\sigma}, D(0,T;\zeta)) \; ,
\label{eq:bs_closed_solution}\end{equation}
where $F^{\text{TVS}}(t,T;\alpha)$ is the TVS forward curve defined by
\begin{equation}
	F^{\text{TVS}}(t,T;\alpha) = I_t \exp\left[\int_t^T\left(r(u)-\ell_{\alpha} \left(u\right)\right)du\right] \; ,
\label{eq:tvs_forward}\end{equation}
while $BS(F,K,T,\sigma,D)$ is the standard BS formula for a European option with forward curve $F$, strike $K$, time to maturity $T$, volatility $\sigma$ and discount factor $D$.

@@

In our work we adopt two learning strategies to compare their performances in terms of overall reward: an \textit{ad hoc} direct policy algorithm and the state of the art proximal policy optimization (PPO) developed in \cite{Schulman2017} and \cite{Schulman2016}.

@@

\subsection{Direct \change{P}olicy \change{A}pproach}\label{sec:direct_policy}
\change{In RL, the learning phase takes place within a set of so-called episodes, which refer to a single run or instance of interaction between the RL agent and an environment, where the agent takes a sequence of actions to achieve a specific goal.}

\change{For our problem, we} \remove{We} consider an episode $\tau$ of length $m+1$ that \remove{takes place on} \change{unfolds over} a discrete time-grid of fixing times expressed in year fractions $\mathcal{T}\coloneqq \{T_0,...,T_k,...,T_m\}$ with $T_0\coloneqq 0$ and $T_m \coloneqq T$ maturity of the option. At a given episodic time $T_k$ the RL agent interacts with the environment: it receives a representation of the environment called state $s_k$ and on the basis of that it selects an action $a_k$ sampling from the current policy $\pi^{\theta_j}$. Here with $\theta_j$ we refer to the set of parameters through which we parameterize the agent policy at the $j$-th algorithm iteration. In our case of study\change{,} the agent\change{, representing the fund manager, has the right to select} \remove{can choose} the composition of the risky asset portfolio, so that the policy is the allocation strategy $\alpha$ introduced in Equation~\eqref{eq:TVS_elegant}:
\begin{equation}
 	a_k = \alpha_{T_k} \quad \forall T_k \in \mathcal{T} \; , \quad a_k \in \mathcal{A}\subset \mathbb{R}^n \; .
\label{eq:action}\end{equation}

Since the value function of the problem depends on the Markovian state $X_t$, the portfolio level $I_t$ and time $t$, our natural choice for the observation state is the following block
\begin{equation}
 	s_{k} \coloneqq \left[X_{T_k}, I_{T_k}, T_k \right] \quad \forall T_k \in \mathcal{T} \; , \quad s_k \in \mathcal{S} \subset \mathbb{R}^{n+2} \; .
\label{eq:state}\end{equation}
In this way, the state contains all the information needed by the agent to take an optimal action, leading to the maximum plain vanilla TVO price.

In this algorithm, we parameterize the agent policy with a feed forward neural network (FFNN), such that $\theta_j$ coincides with the hidden weights, $s_k$ with the inputs, and $a_k$ with the output. In this way, we are dealing with a deterministic policy, where its functional form is given by the neural network. The parameters update is performed as follows: the agent collects a finite batch of experiences interacting with the environment in a set of episodes $\tau$, then the loss function is evaluated as
\begin{equation}
L(\theta_j) = D(0,T;\zeta)\hat{\mathbb{E}}\left[ \Phi(I_T(\alpha(\theta_j))) \right] \; ,
\label{eq:direct_policy_loss}\end{equation}  
where the expectation $\hat{\mathbb{E}}[\dots]$ indicates the empirical average on the batch. Then the parameters are updated by plugging the policy into a stochastic gradient ascent algorithm. The choice of Equation~\eqref{eq:direct_policy_loss} is justified by the fact that, when the algorithm will find the optimal set $\theta^*$, then $L(\theta^*)$ will be a good proxy of the optimal option price.

@@

\subsection{Proximal \change{P}olicy \change{O}ptimization \change{A}pproach}
As for the direct policy approach, we model the pricing problem considering an episode that takes place on the time-grid $\mathcal{T}\coloneqq \{T_0,...,T_k,...,T_m\}$ with $T_0\coloneqq 0$ and $T_m \coloneqq T$. Again we choose as observation state the block defined in Equation~\eqref{eq:state} and the agent policy coincides with the risky allocation strategy $\alpha$ \eqref{eq:action}. An important difference from the previous method is that: once the agent has selected the action $a_k$ sampled from the current policy, it receives at the next time $T_{k+1}$ a reward $r_{k+1}$ generated by the environment.

In our work, we have defined two different reward functions with the purpose to analyze which could help the agent to learn more efficiently the optimal policy $\pi^* = \pi^{\theta^*}$. 
Our first definition is
 \begin{equation}
 	r_{k+1}\change{\coloneqq}\left\{\begin{array}{ll}
 		 D(0,T;\zeta)(I_T(\alpha(\theta_j)) - K)^+ & \text { if } T_{k+1} = T \\
 		0 & \text { otherwise }
 	\end{array}\right.
 \label{eq:reward1} \; . \end{equation}
Therefore, during the whole episode, the agent receives a nil reward except at maturity when the reward coincides with the option intrinsic value. This choice may seem too daring because the agent receives a real feedback of its actions only at the end of the episode, increasing the probability to obtain a slow learning. However, if the agent has learnt $\pi^*$, the average cumulative reward per episode will coincide with the optimal TVO price.

The second reward function we have defined is 
\begin{equation}
 	r_{k+1} \coloneqq  \gamma^{k}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] \; ,
 \label{eq:reward2}\end{equation}
where $\gamma \in[0,1]$ is an hyper-parameter of the PPO, while $V_{\text{BS}}(T_k)$ is a proxy of the residual option price defined by
\begin{equation}
\begin{aligned}
	V_{\text{BS}}(T_k) &\coloneqq BS(F^{\text{TVS}}(T_k,T;\alpha^*_{\text{BS}},K),K,T-T_k,\bar{\sigma},D(T_k,T;\zeta)) \; ,\\
	V_{\text{BS}}(T_0)&=0 \; , 
\end{aligned}
\end{equation} 
with $\alpha^*_{BS}$ the BS optimal strategy \eqref{eq:BS_optimal_strategy} evaluated at the state $s_k$. In this form the agent actions are hidden inside the term $I_t$ used to compute the TVS forward curve $F^{\text{TVS}}$ defined by Equation~\eqref{eq:tvs_forward}. In this case, the reward function does not suffer of nil values for $0<T_k<T$: the RL agent always gets a feedback from the environment for its choices.  \remove{The} \change{Note that the} hyper-parameter $\gamma$ plays the role of a discount factor in the sense that, as $\gamma$ approaches to zero, the RL agent will tend to maximize immediate rewards while neglecting possible larger rewards in the future.
If we take the cumulative reward per episode and set the PPO parameters\footnote{We refer to \cite{Schulman2016} for a more detailed description for the generalized advantage estimation parameter $\lambda$.} $\gamma =\lambda =1$ we obtain
\begin{equation}
	R(\tau) = \sum_{k=0}^{m-1} r_{k+1} \underset{\gamma=\lambda =1}{=} \sum_{k=0}^{m-1}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] = V_{BS}(T) = (I_T-K)^+ \; ,
\end{equation} 
which is equal to the intrinsic value of the option. This result does not depend on the definition of $V_{\text{BS}}$  $\forall T_k<T$, but we conjecture that the closer $V_{\text{BS}}$ is to the value function, the easier the agent is in learning.

@@

\subsection{Local \change{V}olatility \change{D}ynamics}
In this section, we study the TVS control problem assuming a local volatility model for the dynamics of the risky assets. Thus in this case we have a diffusive term in Equation~\eqref{eq:Equity_process} that is a deterministic function both of time and state, i.e. the spot price, $\nu_t=\nu(t,S_t)$. This additional dependency of volatility makes the problem of finding the optimal strategy non-trivial; in fact if we consider the whole securities smiles then the second-order term in the HJB equation \eqref{eq:HJB_general} is not zero and thus a closed formula for $\alpha^*$ is not available anymore. Because of that, one must resort to numerical techniques to recover the problem solution. Unlike the BS model where the strategy depends only on time, in LV dynamics there is no unnecessary information provided by the state block in Equation~\eqref{eq:state} to take the optimal action. Here we consider the same market data ($r_t$, $\mu_t$ and $\nu_t$) as in the Black and Scholes environment to study how the optimal solution changes with the dynamics model.

\begin{figure}[h!]
	\centering
	\includegraphics[width=4in]{LV_direct_policy_approach_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the local volatility \change{scenario}. The solid line with triangles is the learning curve of an agent with free actions, while the solid one is the learning curve of an agent whose actions are parameterized with the maximum drift strategy. The two agents are parameterized by a $20 \times 10 \times 5$ FFNN with tanh activation function. The loss function is computed according to Equation~\eqref{eq:direct_policy_loss}. The optimizer adopted is the RMSprop with a learning rate of $10^{-3}$ for the solid line with triangles while $10^{-4}$ for the solid one. The horizontal grey-dashed lines delimit the 99\% confidence interval of the MC price obtained by maximizing path-wise the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_lv_direct_policy}
\end{figure}  
Our first way to tackle the problem is to exploit the results obtained in the BS case. Thus we train by direct policy algorithm a deep $20\times 10\times 5$ neural network with tanh activation function and a RMSprop optimizer. We report in Figure~\ref{fig:result_lv_direct_policy} the corresponding learning curve (solid line with triangles). From this, we observe that the network has learnt some good policy since the curve grows as the number of training epochs increases until it saturates at a certain value. To try to measure the performance of the policy selected by the agent, we can build a n\"aif strategy called \quotations{baseline}. By looking at the Markovian projection in \eqref{eq:markovian_projection}-\eqref{eq:local_drift} and the first term in the HJB \eqref{eq:HJB_general}, a natural choice for the baseline is to maximize the TVS local drift. In other words this coincides by applying the path-wise the Black and Scholes solution of Equation~\eqref{eq:BS_optimal_strategy}; since here we deal with a free allocation strategy, we can simply use our analytical result \eqref{eq:analytical_solution_free_strategy}. In the same Figure~\ref{fig:result_lv_direct_policy} we report the 99\% confidence interval of the MC price obtained with the maximum drift baseline as two red-dashed lines. As we can see, the optimized loss function is compatible with the baseline price. Following the theoretical result of the HJB equation, we can assert that the agent has learnt a sub-optimal policy. Since we LV model differs from the BS one for a corrective term in the HJB \eqref{eq:HJB_general}, we expect that the optimal solution  in the LV framework will be in a close region of the maximum drift strategy. Thus we train another agent with the direct policy learning by parameterizing its action with the baseline strategy and choosing a smaller learning rate of $10^{-4}$. With this parameterization, at each observational time $T_k\in\mathcal{T}$ the risky allocation strategy is obtained by summing the network output with the Equation~\eqref{eq:analytical_solution_free_strategy}. Even in this case, the corresponding learning curve (solid black line in Figure~\ref{fig:result_lv_direct_policy}) is stuck in the maximum drift strategy. The first possible reason for this behaviour is that the learning strategy of the direct policy approach is too simple for the learning task. The second interpretation is that the volatility versor in the HJB does not affect significantly the optimum location since it is a second-order term and is lost in the MC error.

>>>>

\usepackage[]{latexsym}
\usepackage{graphicx}
\usepackage{xcolor, cancel}

@@

\usepackage[normalem]{ulem}
\newcommand\soutpars[1]{\let\helpcmd\sout\parhelp#1\par\relax\relax}
\long\def\parhelp#1\par#2\relax{%
  \helpcmd{#1}\ifx\relax#2\else\par\parhelp#2\relax\fi%
}
\newcommand{\change}[1]{{\color{red} {#1}}}%{{#1}}
\newcommand{\remove}[1]{{\color{red} \soutpars{{#1}}}}%{}
\newcommand{\removeeq}[1]{{\color{red} \sout{{#1}}}}%{}
\newcommand{\changeB}[1]{{\color{blue} {#1}}}%{{#1}}
\newcommand{\removeB}[1]{{\color{blue} \soutpars{{#1}}}}%{}
\newcommand\hcancel[2][red]{\setbox0=\hbox{$#2$}%
\rlap{\raisebox{.45\ht0}{\textcolor{#1}{\rule{\wd0}{1pt}}}}#2}

@@

In this way, the TVS is only defined in the two contracts client-fund and fund-\change{option writer} \remove{fund-bank}. The fund manager is not implementing the strategy by trading in the market, and \change{s}he  is not subject to additional costs to access the market. On the other way, the \remove{bank} \change{writer} is paying such costs since she is actively hedging the call option sold to the manager.  The \remove{bank} \change{writer} trading activity implemented to actively hedge the option requires funding the collateral procedures of the hedging instruments along with any lending/borrowing fee.
We remark that\change{, upon entering into the contract,}  the choices of the manager trading activity \remove{are} \change{can be seen as} stochastic processes since they \change{are not known {\it a priori} and }will depend on the market evolution. Thus, neither the manager strategy nor the \remove{bank} \change{writer} financing costs for hedging the option can be written in the fund-\remove{bank}\change{writer} contract\change{: the only contractual elements are the protection strike and maturity, the trading market of the manager and the target volatility of the TVS portfolio}. For this reason, the price of a financial product sold by the \remove{bank} \change{writer} is adjusted to include any valuation adjustment due to the trading activity.

Our aim is to find the most expensive investing strategy from the point of view of financing costs for the \remove{bank} \change{writer} that the manager could choose in the market. In other words, we want to determine the worst-case scenario for the \remove{bank} \change{option-seller}. In the current section, we proceed by defining the price process of the TVS so that we can highlight the impact of valuation adjustments.

@@

Then, we introduce the strategy price process $V_t$, and we define the deflated gain process $\Bar{G}_t^V$ as \remove{given by}
\begin{equation}
    \Bar{G}_t^V \coloneqq \frac{V_t}{B_t} + \int_0^t\frac{V_u\phi_u}{B_u}du \; ,
\end{equation}
where $\phi_t$ are the running fees earned by the fund manager for \remove{his} \change{her} activity. We assume that the strategy is self-financing, so that we can write
\begin{equation}
    d\Bar{G}_t^V = q_t \cdot d\bar{G}_t \; ,
\label{eq:self_financing}\end{equation}
where $q_t^i$ is the quantity invested in the $i$-th security.\footnote{In all formulae we use dot notation for scalar product between vectors, i.e. $a \cdot b = \sum_i a_i b_i$, or between matrix and vector, i.e. $A \cdot b = \sum_j a_{ij}b_j$ or $b \cdot A = \sum_i b_i a_{ij}$.}

@@

\begin{proposition}
When the underlying securities follow a Black and Scholes \change{QUA IO LASCEREI MODEL (REFEREE PROPONE SCENARIO)} model with deterministic rates, the optimal strategy for a non-decreasing European payoff consists in minimizing the local drift function, independently of the current state $I_t$
\begin{equation}
    \alpha^*(t) \coloneqq \argmin_\alpha \frac{\alpha \cdot \mu(t)}{\|\alpha \cdot \nu(t) \|} \; .
\label{eq:BS_optimal_strategy}
\end{equation}
Analogously, the optimal strategy for a non-increasing European payoff consists in maximizing the local drift function:
\begin{equation}
    \alpha^*(t) \coloneqq \argmax_\alpha \frac{\alpha \cdot \mu(t)}{\norm{\alpha \cdot \nu(t)}} \; .
\label{eq:BS_optimal_strategy_put}
\end{equation}
\end{proposition}
The absence of stochastic elements in Equation~\eqref{eq:BS_optimal_strategy} makes the optimal strategy known \textit{a priori} with no \change{numerical} simulation needed \change{to solve the control problem}; in fact \remove{one can solve the optimization problem once for all $\forall t \in \mathcal{T}$ just looking at the market data $\mu(t)$ and $\nu(t)$ for the securities} \change{, once the risky-assets market data $\mu(t)$ and $\nu(t)$, it is sufficient to apply Equation~\eqref{eq:BS_optimal_strategy} $\forall t \in \mathcal{T}$  to retrieve the deterministic optimal strategy}. Once $\alpha^*$ is known, then one can price the payoff by the following BS formula
\begin{equation}
	V_0^{\text{BS}} = BS(F^{\text{TVS}}(0,T;\alpha^*), K, T, \bar{\sigma}, D(0,T;\zeta)) \; ,
\label{eq:bs_closed_solution}\end{equation}
where $F^{\text{TVS}}(t,T;\alpha)$ is the TVS forward curve defined by
\begin{equation}
	F^{\text{TVS}}(t,T;\alpha) = I_t \exp\left[\int_t^T\left(r(u)-\ell_{\alpha} \left(u\right)\right)du\right] \; ,
\label{eq:tvs_forward}\end{equation}
while $BS(F,K,T,\sigma,D)$ is the standard BS formula for a European option with forward curve $F$, strike $K$, time to maturity $T$, volatility $\sigma$ and discount factor $D$.

@@

In our work we adopt two learning strategies to compare their performances in terms of overall reward: an \textit{ad hoc} direct policy \change{search} algorithm and the state of the art proximal policy optimization (PPO) developed in \cite{Schulman2017} and \cite{Schulman2016}.

@@

\subsection{Direct \change{P}olicy \change{Search} \change{A}pproach}\label{sec:direct_policy}
\change{In RL, the learning phase takes place within a set of so-called episodes, which refer to a single run or instance of interaction between the RL agent and an environment, where the agent takes a sequence of actions to achieve a specific goal.}

\change{For our problem, we} \remove{We} consider an episode $\tau$ of length $m+1$ that \remove{takes place on} \change{unfolds over} a discrete time-grid of fixing times expressed in year fractions $\mathcal{T}\coloneqq \{T_0,...,T_k,...,T_m\}$ with $T_0\coloneqq 0$ and $T_m \coloneqq T$ maturity of the option. At a given episodic time $T_k$ the RL agent interacts with the environment: it receives a representation of the environment called state $s_k$ and on the basis of that it selects an action $a_k$ sampling from the current policy $\pi^{\theta_j}$. Here with $\theta_j$ we refer to the set of parameters through which we parameterize the agent policy at the $j$-th algorithm iteration. In our case of study\change{,} the \change{RL} agent\change{, representing the fund manager, has the right to select} \remove{can choose} the composition of the risky asset portfolio, so that the policy is the allocation strategy $\alpha$ introduced in Equation~\eqref{eq:TVS_elegant}:
\begin{equation}
 	a_k = \alpha_{T_k} \quad \forall T_k \in \mathcal{T} \; , \quad a_k \in \mathcal{A}\subset \mathbb{R}^n \; .
\label{eq:action}\end{equation}

Since the value function of the problem depends on the Markovian state $X_t$, the portfolio level $I_t$ and time $t$, our natural choice for the observation state is the following block
\begin{equation}
 	s_{k} \coloneqq \left[X_{T_k}, I_{T_k}, T_k \right] \quad \forall T_k \in \mathcal{T} \; , \quad s_k \in \mathcal{S} \subset \mathbb{R}^{n+2} \; .
\label{eq:state}\end{equation}
In this way, the state contains all the information needed by the agent to \remove{take an} \change{select the} optimal action, leading to the maximum plain vanilla TVO price.

\change{Once the agent has selected the action $a_k$ sampled from the current policy, it receives at the next time $T_{k+1}$ a reward $r_{k+1}$ generated by the environment. In this algorithm we have defined the reward function as follows}
 \begin{equation}
 	\change{r_{k+1}\change{\coloneqq}\left\{\begin{array}{ll}
 		 D(0,T;\zeta)(I_T(\alpha(\theta_j)) - K)^+ & \text { if } T_{k+1} = T \\
 		0 & \text { otherwise }
 	\end{array}\right.
 \label{eq:reward1} \; .} \end{equation}

 \change{Therefore, during the whole episode, the agent receives a nil reward except at maturity when the reward coincides with the option intrinsic value. This choice may seem too daring because the agent receives a real feedback of its actions only at the end of the episode, increasing the probability to obtain a slow learning. However, if the agent has learnt the optimal policy $\pi^* = \pi^{\theta^*}$, the average cumulative reward per episode will coincide with the optimal TVO price.}

In this algorithm, we parameterize the agent policy with a feed forward neural network (FFNN), such that $\theta_j$ coincides with the hidden weights, $s_k$ with the \remove{inputs} \change{input neurons}, and $a_k$ with the output \change{ones}. In this way, we are dealing with a deterministic policy, \change{that maps each state directly to a single action without any randomness or probability distribution;} \remove{where its} \change{the map} functional form is given by the neural network. The parameters update is performed as follows: the agent collects a finite batch of experiences interacting with the environment in a set of episodes $\tau$, then the loss-\change{value} function is evaluated as
\begin{equation}
\change{L(\theta_j) =\hat{\mathbb{E}}\left[ \sum_{k=0}^{m-1} r_{k+1} | \theta_j \right]} \; ,
\label{eq:direct_policy_loss}\end{equation}
\begin{equation}
\hcancel[red]{\change{L(\theta_j) = D(0,T;\zeta)\hat{\mathbb{E}}\left[ \Phi(I_T(\alpha(\theta_j))) \right]}} \; ,
\end{equation}  
where the expectation $\hat{\mathbb{E}}[\dots]$ indicates the empirical average on the batch. Then the parameters are updated by plugging the policy into a stochastic gradient ascent algorithm \change{with the aim of maximizing Equation~\eqref{eq:direct_policy_loss}}. \remove{The choice of Equation~\eqref{eq:direct_policy_loss} is justified by the fact that, when the algorithm will find the optimal set $\theta^*$, then $L(\theta^*)$ will be a good proxy of the optimal option price.}

@@

\subsection{Proximal \change{P}olicy \change{O}ptimization \change{A}pproach}
As for the direct policy \change{search} approach, we model the pricing problem considering an episode that takes place on the time-grid $\mathcal{T}\coloneqq \{T_0,...,T_k,...,T_m\}$ with $T_0\coloneqq 0$ and $T_m \coloneqq T$. Again we choose as observation state the block defined in Equation~\eqref{eq:state} and the agent policy coincides with the risky allocation strategy $\alpha$ \eqref{eq:action}. \remove{An important difference from the previous method is that: once the agent has selected the action $a_k$ sampled from the current policy, it receives at the next time $T_{k+1}$ a reward $r_{k+1}$ generated by the environment.}

\remove{In our work, we have defined two different reward functions with the purpose to analyze which could help the agent to learn more efficiently the optimal policy $\pi^* = \pi^{\theta^*}$. 
Our first definition is} 
 \begin{equation}
 	\hcancel[red]{\change{r_{k+1} =\left\{\begin{array}{ll}
 		 D(0,T;\zeta)(I_T(\alpha(\theta_j)) - K)^+ & \text { if } T_{k+1} = T \\
 		0 & \text { otherwise }
 	\end{array}\right.}}
  \; . \end{equation}
\remove{Therefore, during the whole episode, the agent receives a nil reward except at maturity when the reward coincides with the option intrinsic value. This choice may seem too daring because the agent receives a real feedback of its actions only at the end of the episode, increasing the probability to obtain a slow learning. However, if the agent has learnt $\pi^*$, the average cumulative reward per episode will coincide with the optimal TVO price.}

\change{As the PPO algorithm is more sophisticated than the previous one, we have decided to test its performance by means of two different reward functions. The former is the same as defined by Equation~\eqref{eq:reward1}, while the latter is given by}
\remove{The second reward function we have defined is}
\begin{equation}
 	r_{k+1} \coloneqq  \gamma^{k}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] \; ,
 \label{eq:reward2}\end{equation}
where $\gamma \in[0,1]$ is an hyper-parameter of the PPO, while $V_{\text{BS}}(T_k)$ is a proxy of the residual option price defined by
\begin{equation}
\begin{aligned}
	V_{\text{BS}}(T_k) &\coloneqq BS(F^{\text{TVS}}(T_k,T;\alpha^*_{\text{BS}},K),K,T-T_k,\bar{\sigma},D(T_k,T;\zeta)) \; ,\\
	V_{\text{BS}}(T_0)&=0 \; , 
\end{aligned}
\end{equation} 
with $\alpha^*_{BS}$ the BS optimal strategy \eqref{eq:BS_optimal_strategy} evaluated at the state $s_k$. In this form the agent actions are hidden inside the term $I_t$ used to compute the TVS forward curve $F^{\text{TVS}}$ defined by Equation~\eqref{eq:tvs_forward}. In this case, the reward function does not suffer of nil values for $0<T_k<T$: the RL agent always gets a feedback from the environment for its choices.  \remove{The} \change{Note that the} hyper-parameter $\gamma$ plays the role of a discount factor in the sense that, as $\gamma$ approaches to zero, the RL agent will tend to maximize immediate rewards while neglecting possible larger rewards in the future.
If we take the cumulative reward per episode and set the PPO parameters\footnote{We refer to \cite{Schulman2016} for a more detailed description for the generalized advantage estimation parameter $\lambda$.} $\gamma =\lambda =1$ we obtain
\begin{equation}
	R(\tau) = \sum_{k=0}^{m-1} r_{k+1} \underset{\gamma=\lambda =1}{=} \sum_{k=0}^{m-1}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] = V_{BS}(T) = (I_T-K)^+ \; ,
\end{equation} 
which is equal to the intrinsic value of the option. This result does not depend on the definition of $V_{\text{BS}}$  $\forall T_k<T$, but we conjecture that the closer $V_{\text{BS}}$ is to the value function, the easier the agent is in learning.

@@

\subsection{Local \change{V}olatility \change{D}ynamics}
In this section, we study the TVS control problem assuming a local volatility model for the dynamics of the risky assets. Thus in this case we have a diffusive term in Equation~\eqref{eq:Equity_process} that is a deterministic function both of time and state, i.e. the spot price, $\nu_t=\nu(t,S_t)$. This additional dependency of volatility makes the problem of finding the optimal strategy non-trivial\change{, as a closed form formula cannot be derived under this scenario}; in fact if we consider the whole \remove{securities} \change{volatility} smiles \change{of the securities,} then the second-order term in the HJB equation \eqref{eq:HJB_general} \remove{is not zero and thus a closed formula for $\alpha^*$ is not available anymore} \change{does not elide and consequently the only way to derive $\alpha^*$ is through a numerical method}. \remove{Because of that, one must resort to numerical techniques to recover the problem solution.} \remove{Unlike the BS model where the strategy depends only on time, in LV dynamics there is no unnecessary information provided by the state block in Equation~\eqref{eq:state} to take the optimal action} \change{The additional dependence on the state $S_t$ under LV dynamics will cause the agent to consider also the $X_t$ component value in the  state block \eqref{eq:state}, unlike under the BS scenario where this information is redundant as the optimal strategy was function only of time \eqref{eq:TVS_last}}. Here we consider the same market data ($r_t$, $\mu_t$ and $\nu_t$) as in the Black and Scholes environment to study how the optimal solution changes with the dynamics model.

\begin{figure}[h!]
	\centering
	\includegraphics[width=4in]{LV_direct_policy_approach_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the local volatility \change{scenario}. The solid line with triangles is the learning curve of an agent with free actions, while the solid one is the learning curve of an agent whose actions are parameterized with the maximum drift strategy. The two agents are parameterized by a $20 \times 10 \times 5$ FFNN with tanh activation function. The loss function is computed according to Equation~\eqref{eq:direct_policy_loss}. The optimizer adopted is the RMSprop with a learning rate of $10^{-3}$ for the solid line with triangles while $10^{-4}$ for the solid one. The horizontal grey-dashed lines delimit the 99\% confidence interval of the MC price obtained by maximizing path-wise the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_lv_direct_policy}
\end{figure}  
Our first way to tackle the problem is to exploit the results obtained in the BS case. Thus we train by direct policy algorithm a deep $20\times 10\times 5$ neural network with tanh activation function and a RMSprop optimizer. We report in Figure~\ref{fig:result_lv_direct_policy} the corresponding learning curve (solid line with triangles). From this, we observe that the network has learnt some good policy since the curve grows as the number of training epochs increases until it saturates at a certain value. To try to measure the performance of the policy selected by the agent, we can build a n\"aif strategy called \quotations{baseline}. By looking at the Markovian projection in \eqref{eq:markovian_projection}-\eqref{eq:local_drift} and the first term in the HJB \change{equation} \eqref{eq:HJB_general}, \change{we see that the control problem is similar to the BS one with an additional second order term. Thus} a natural choice for the baseline is to maximize the TVS local drift \remove{. In other words this coincides} by applying the path-wise the Black and Scholes solution of Equation~\eqref{eq:BS_optimal_strategy}. \change{S}ince here we deal with a free allocation strategy, we can simply use our analytical result \eqref{eq:analytical_solution_free_strategy}. In the same Figure~\ref{fig:result_lv_direct_policy} we report the 99\% confidence interval of the MC price obtained with the maximum drift baseline as two red-dashed lines. As we can see, the optimized loss function is compatible with the baseline price. Following the theoretical result of the HJB equation, we can assert that the agent has learnt a sub-optimal policy. Since we LV model differs from the BS one for a corrective term in the HJB \change{equation} \eqref{eq:HJB_general}, we expect that the optimal solution  in the LV framework will be in a close region of the maximum drift strategy. Thus we train another agent with the direct policy learning by parameterizing its action with the baseline strategy and choosing a smaller learning rate of $10^{-4}$. With this parameterization, \change{the RL agent actions are computed by summing} at each observational time $T_k\in\mathcal{T}$ \remove{the risky allocation strategy is obtained by summing} the FFNN output with the Equation~\eqref{eq:analytical_solution_free_strategy}\change{, giving rise to the manager's risky allocation strategy}. Even in this case, the corresponding learning curve (solid black line in Figure~\ref{fig:result_lv_direct_policy}) is stuck in the maximum drift strategy. The first possible reason for this behaviour is that the learning strategy of the direct policy approach is too simple for the learning task. The second interpretation is that the volatility versor in the HJB does not affect significantly the optimum location since it is a second-order term and is lost in the MC error.
